---
title: "Delfina take-home assessment"
author: "Leonardo Hernandez Espinoza"
date: "2022-07-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(data.table)
library(formattable)
```

## Import the data

```{r}
# If you are testing this outside my repo, please change the url
data_url <- "~/R/delfina-ds-interview/data/birthdata2020.csv"
birth_data <- read.csv(data_url, header = TRUE)
```

## Task 1:  What is the proportion of births by gestational age at delivery
I am answering this question as the proportion of births at each one of the combined gestation weeks based on the column COMBGEST.

Given that COMBGEST uses the value 99 as a filler value to denote missing data I will remove that value. 
I am also filtering the data to limit the results to values between 20 and 50 weeks, that range was selected based on information found [here](https://www.health.ny.gov/community/pregnancy/why_is_40_weeks_so_important.htm#:~:text=How%20long%20is%20full%20term%3F).

```{r}
birth_data %>% 
  filter(COMBGEST %in% (20:50)) %>% 
  ggplot(aes(COMBGEST))+
  geom_bar(aes(y = (..count..)/sum(..count..)), fill = 'blue') +
  theme_classic()+
  xlab("Number of weeks at birth")+
  ylab("Proportion of births")
```
<br>
As we can see, the largest proportion of births occur between 39 and 41 weeks of gestation.


## Task 2:  What is the prevalence of preterm birth in this dataset?
To answer this question I will create a flag variable (is_preterm) to indicate if the birth occurred prior to 37 weeks of gestation (based on COMBGEST). In order to better calculate the proportion of preterm birth that occurred during 2020, I will exclude those cases where <br>COMBGEST =99.

From this point on, the is_preterm column will be used multiple times so I will fist add it to the data set.
```{r}
birth_data <- birth_data %>% 
  filter(COMBGEST != 99) %>% 
  mutate(is_preterm = case_when(
    COMBGEST < 37 ~ "Yes",
    TRUE ~ "No"
  ))


```
With the new column added, I then evaluate the proportions of preterm births
```{r}
 task2_table <- birth_data %>% 
  group_by(is_preterm) %>% 
  summarise(n_cases = n()) %>%
  mutate(percentage = 100* round(n_cases / sum(n_cases), 4)) %>% 
  data.table()
```
I will use the formattable package for better aesthetics 
```{r}
formattable(task2_table)
```

We can see that from all the observations in the data set, less ~12% of the cases are preterm births.

## Tast 3: Summarizes the relationship between preterm birth and the following four variables:

  - Maternal age (MAGER)

  - Maternal race/Hispanic origin (MRACEHISP)

  - Previous preterm birth (RF_PPTERM)

  - Total number of prenatal visits (PREVIS)


Based on the example given, 
